---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bskernel

<!-- badges: start -->
![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)
![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)

<!-- badges: end -->

This package provides flexible auto-covariance kernel construction via inverse Fourier transforms of B-spline bases, as well as some handy tools for optimisation. Please note, if you have found yourself in the GitHub repo, please navigate [here](https://astfalckl.github.io/bskernel/) for the pkgdown page.

## Installation

You can install the development version of `bskernel` from GitHub with

```r
devtools::install_github("astfalckl/bskernel")
```

## Quickstart

The `bskernel` package provides tools for constructing auto-covariance functions via the inverse Fourier transform of B-spline basis functions on the spectral domain. The workhorse functions are 

  - `build_bspline_design_matrix`
  - `reconstruct_acf`

Below is a minimal example that recreates Figure 1 of the paper.

```{r}
library(bskernel)

k <- 1 
knots <- seq(-0.125, 0.5, 0.125)
omega <- seq(0, 0.5, length.out = 100)
tau <- seq(0, 7.5, length = 100)

spline_bases <- build_bspline_design_matrix(omega, knots, k)

acfs <- sapply(1:4, function(i){
  idx <- rep(0, 4)
  idx[i] <- 1
  Re(reconstruct_acf(idx, knots, k, tau))
})
```

Note that `reconstruct_acf()` is designed to summate over a bases expansion, here we iteratively select the basis member. The spline and acf bases are plotted below (plotting code is suppressed, see `.Rmd` file).

```{r, echo = FALSE, fig.height=3}
library(ggplot2)
library(patchwork)
library(tidyr)

basis_df <- as.data.frame(spline_bases)
basis_df$omega <- omega

basis_long <- pivot_longer(
  basis_df,
  cols = -omega,
  names_to = "basis",
  values_to = "value"
)

acf_df <- as.data.frame(acfs)
acf_df$tau <- tau

acf_long <- pivot_longer(
  acf_df,
  cols = -tau,
  names_to = "basis",
  values_to = "value"
)

plot_spline <- ggplot(basis_long, aes(x = omega, y = value, group = basis, linetype = basis)) +
  geom_line() +
  labs(
    x = expression(omega),
    y = "Spectral Density",
    title = "Spline Bases"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

plot_acf <- ggplot(
  acf_long,
  aes(x = tau, y = value, group = basis, linetype = basis)
) +
  geom_line() +
  labs(
    x = expression(tau),
    y = "Auto-covariance",
    title = "ACF Bases"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

plot_spline + plot_acf
```

## Articles

I have written a number of tutorials:

 - [Function reference](reference/index.html)

### Citation

This package accompanies the paper

LC Astfalck. (2025). *''Universal approximation of auto-covariance functions via spline kernels''*. Submitted to the Journal of the Royal Statistical Society: Series B. [[Available here.]](https://arxiv.org/abs/2506.21953)